### 2019-10-29 main_text/main_text.tex
MM edits to main text

	Through physical and mathematical modeling, we investigate a more complex
	switch system where the bistability is due, as we will show, to a
	membrane transport protein. Such a switch is common for metabolic processes in biology, for reasons discussed below. So far, models we have found are either very extensive descriptions of specific, complicated systems, or they are very basic. To enable an intuitive and fundamental understanding of the behavior, we develop a simple model of a minimal system that only has the essential components and interactions for the questions we pose, while still modeling every component explicitly and discussing the necessary model complexity for a physically correct model.

@@

	For our modeling, we focus on the \emph{xapABR} genetic circuit from
	\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
	complex. Instead of lactose, its purpose is to make use of the nucleoside
	xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
	circuit is made up of two operons: one that encodes for XapR and another
	that encodes for XapA and XapB. XapR appears to be a transcription factor
	that is induced by xanthosine and activates the \emph{xapAB} promoter,
	in close analogy to AraC, XylR, and also LacI, although the latter acts via repression.
	The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
	but the promoter architecture and function is not yet fully understood. The
	transcription of \emph{xapR} seems to be constitutive and not
	auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
	factors suggests that XapR appears in dimers where one dimer can bind two
	xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
	nucleoside phosphorylase that degrades xanthosine into components (ribose
	and xanthine) that can be fed into metabolic
	pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
	membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapAB} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine are nonspecifically transported\todo{``permeate'' wouldn't include active transport (NupC \& NupG), would it?} into and out of the cell (discussed in more detail below).
		\item XapR is induced by xanthosine.
		\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA \& XapB is balanced by decay or dilution through cell divisions.
	\end{itemize}
	Because xanthosine induces the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

	\begin{figure}[h!]
		\centering
		\includegraphics{media/XapSystem.eps}
		\caption{{\bf Model of the \emph{xapAB} circuit.}
			The XapR dimers are induced by xanthosine and the induced XapR binds
			cooperatively as an activator to the \emph{xapAB} promoter. For
			these two steps, quasi-equilibrium is assumed. If both XapR binding
			sites are occupied and the polymerase is bound, the gene is
			transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
			$\gamma_{\n{m}}$, and both proteins are translated at rate
			$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
			Similarly, XapB imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
			k_{\n{nup}}$, respectively.}
		\label{fig2:model}
	\end{figure}

@@

	There are two significant ways of transport of xanthosine across the cell membrane. In the induced system, the main transporter is XapB, whereas in the uninduced system, there is almost no XapB. Instead, xanthosine can enter the cell through the two nucleoside transporters NupC and NupG, which have a very low affinity for xanthosine~\cite{Norholm2001}. All these transporters, XapB, NupC, and NupG, are powered by the proton gradient across the membrane~\cite{Norholm2001}, which is why we assume their kinetic scheme to be similar to that of the \emph{lac} permease (as it is described in~\cite{Kaback2015}). There can be import and export of xanthosine, and which one dominates depends on the proton and xanthosine concentrations on the two sides of the membrane. In both cases, a proton and a substrate need to bind to the transporter on one side of the membrane and detach from it on the other side before the empty transporter moves back to the other side again. We refer the reader to~\nameref{S1_Text} for a detailed description of the transport and its modeling, and just state the result here. We model influx and efflux separately. For XapB, we use Michaelis-Menten kinetics with parameters $k_{\n{b, i}}$, $K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux. For the Nup transporters, we can linearize because of their low affinity for xanthosine and are left with the rate parameters $k_{\n{nup}}$ for influx and $\xi k_{\n{nup}}$ for efflux.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/VaryingC.eps}
		\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
			All parameters but $\n{[c]_a}$ are as presented in
			Table~\ref{table1:nondim}. The extracellular xanthosine
			concentration in these plots is $\n{[c]_a}=10^{1}$ in (A) and
			$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall that $\n{[c]_{a}} \defeq
			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$ is dimensionless).
			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline) up
			or down, the blue curve (mRNA nullcline) is unchanged (see
			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
			only the lower fixed point (fixed point number 1), whereas in (B)
			there is only the upper one (fixed point number 3). In between lies
			the bistable case that was shown in Fig~\ref{fig4:bistability}.}
		\label{fig5:extraxanth}
	\end{figure}

@@

	\paragraph*{The roles of XapA and XapB.}
	While it is clear that the bistability in the model system is due to the
	feedback loop from XapA and XapB, it is not obvious if both XapA and XapB
	are necessary. It turns out that the bistability is due to XapB only. XapA
	is neither sufficient nor necessary and, within the estimated parameter
	regime, does not even have a significant influence on the system. This can
	be seen from the plots in figure~\ref{fig6:xapAB}. The degradatory effect of XapA lowers the xanthosine and protein concentration at the upper
	fixed point by a small amount and could, in principle, thereby make the
	high-expression solution vanish. For our choice of all other parameters,
	it only vanishes for $k_{\n{\alpha}} > 10^4$ which is far from what has been
	measured. However, a higher effective rate could, in principle, be achieved by
	different translation rates of XapA and XapB (see simplifications of the
	model in~\nameref{S1_Text}). Hence, we cannot exclude the possibility that
	XapA becomes so strong that it makes bistability impossible, but this is an
	extreme case. XapB, on the other hand, is essential; without it the system
	only has the one fixed point at low expression.\footnote{
		Seeger et.\ al.\cite{Seeger1995} observed that $\Delta xapB$ mutants could
		survive, but grew extremely slowly, with xanthosine as the only carbon
		source, which makes sense in light of Fig~\ref{fig6:xapAB}.
		With \textit{xapB} removed, the switch never activates and the cells are
		forced to survive with an extremely meager quantity of XapA to metabolize
		the abundant xanthosine.}

@@

	For a cell, the insignificance of the effect of XapA is a useful feature: by coupling XapA and XapB on an
	operon, XapA is switched on and off together with XapB but it does not
	significantly disturb this adaptation mechanism, while its kinetic
	parameters and expression levels can be chosen 
	somewhat freely as necessary for metabolism. By
	having a membrane transporter gene on an operon whose expression is
	activated by the transporter substrate, the expression of a whole set of
	enzymes can be turned on and off depending on the presence of the substrate.
	It seems likely that this mechanism of short-term adaptation of a single cell to
	its environment may be used by cells for many metabolic processes.

@@

	If, as a purely theoretical consideration, we remove either the second
	xanthosine binding site on XapR or the second XapR binding site on the
	promoter, leaving cooperativity in one place of the system, we find that the system still has a bistable parameter regime.
	However, this bistable parameter range is only narrow, which makes the system
	less stable: small stochastic fluctuations in the parameter values can collapse the system to monostability, possibly leaving it in the wrong state and without its ability to adapt. But only when the second binding site is removed in both places, leaving no cooperativity in the system, the
	equations are insufficiently non-linear to produce bistability. An example
	of the three scenarios (only cooperative XapR, only cooperative promoter, no cooperativity) can be seen in Fig~\ref{fig7:coop}. It follows that
	there need to be either two xanthosine binding sites on XapR or two XapR
	binding sites on the promoter (or both) in order to obtain a switch-like
	behavior.

@@

	One can also ask how much cooperative interaction is needed between the two
	binding sites. For the promoter, the amount of cooperativity is given by $\Delta E_{\n{coop}}$ in
	our model, and we find that setting $\Delta E_{\n{coop}} = 0$ has almost no
	influence on the phase diagrams. For XapR, we cannot test how much interaction is needed: the two binding sites interact
	indirectly, because the active state is much likelier if two xanthosine
	molecules are bound, and thus there is no tuning parameter for the cooperative interaction
	like $\Delta E_{\n{coop}}$ in the case of the promoter.

@@

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state, because they do not show the barrier height between the two states. In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations.

@@

>>>>

	Through physical and mathematical modeling, we investigate a more complex
	switch system where the bistability is due, as we will show, to a
	membrane transport protein. Such a switch is common for
	metabolic processes in biology, for reasons discussed below.
	Existing models in the literature tend towards one of two extremes:
	either highly detailed descriptions of specific, complicated systems (e.g.,~\cite{Wong1997}),
	or Hill function descriptions that coarse-grain all complexity
	into a few parameters with inscrutable microscopic physical meaning.
	We aim for a middle ground in this work.
	We seek an intuitive understanding through a simple model of a minimal system,
	with only the essential components and interactions for the questions we pose.
	Yet we still model these components explicitly and
	discuss the necessary model complexity for a physically correct model.

@@

	For our modeling, we focus on the \emph{xapABR} genetic circuit from
	\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
	complex. Instead of lactose, its purpose is to make use of the nucleoside
	xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
	circuit is made up of two operons: one that encodes for XapR and another
	that encodes for XapA and XapB. XapR appears to be a transcription factor
	that is induced by xanthosine and activates the \emph{xapAB} promoter,
	in close analogy to AraC, XylR, and also LacI.
	\footnote{One might object that LacI represses its target operon, while
	XapR, AraC, and XylR activate their target operons. However, the analogy
	we wish to draw is that the qualitative logic of their inducers are all
	identical, i.e., the presence of their respective inducer causes their
	target operon to be transcribed.}
	The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
	but the promoter architecture and function is not yet fully understood. The
	transcription of \emph{xapR} seems to be constitutive and not
	auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
	factors suggests that XapR appears in dimers where one dimer can bind two
	xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
	nucleoside phosphorylase that degrades xanthosine into components (ribose
	and xanthine) that can be fed into metabolic
	pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
	membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapAB} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine permeate\todo{``permeate'' wouldn't include active transport (NupC \& NupG), would it? MM: I think it's sufficiently vague to capture all this, details are too subtle to worry about here} into and out of the cell (discussed in more detail below).
		\item XapR is induced by xanthosine.
		\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA \& XapB is balanced by decay or dilution through cell divisions.
	\end{itemize}
	Because xanthosine induces the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

	\begin{figure}[h!]
		\centering
		\includegraphics{media/XapSystem.eps}
		\caption{{\bf Model of the \emph{xapAB} circuit.}
			The XapR dimers are induced by xanthosine and the induced XapR binds
			cooperatively as an activator to the \emph{xapAB} promoter. For
			these two steps, quasi-equilibrium is assumed. If both XapR binding
			sites are occupied and the polymerase is bound, the gene is
			transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
			$\gamma_{\n{m}}$, and both proteins are translated at rate
			$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
			Similarly, XapB is treated as a Michaelis-Menten enzyme which imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
			k_{\n{nup}}$, respectively.}
		\label{fig2:model}
	\end{figure}

@@

	\todo{MM: do these paragraphs on xanthosine dynamics have an appropriate level of detail now, or too much, or not enough?}
	There are two significant mechanisms for transport of xanthosine across
	the cell membrane. In the induced system, the main transporter is XapB,
	whereas in the uninduced system, there is almost no XapB. Instead,
	xanthosine can enter the cell through the two nucleoside transporters
	NupC and NupG, which have a very low affinity for
	xanthosine~\cite{Norholm2001}. All these transporters, XapB, NupC, and
	NupG, are powered by the proton gradient across the
	membrane~\cite{Norholm2001}, which is why we assume their kinetic scheme
	to be similar to that of the \emph{lac} permease (as it is described
	in~\cite{Kaback2015}). There can be import and export of xanthosine, and
	which one dominates depends on the proton and xanthosine concentrations
	on the two sides of the membrane. In both cases, a proton and a
	substrate need to bind to the transporter on one side of the membrane
	and detach from it on the other side before the empty transporter moves
	back to the other side again. We refer the reader to~\nameref{S1_Text}
	for a detailed description of the transport and its modeling, and just
	state the result here. We model influx and efflux separately. For XapB,
	we use Michaelis-Menten kinetics with parameters $k_{\n{b, i}}$,
	$K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux.
	For the Nup transporters, we can linearize because of their low affinity
	for xanthosine and are left with the rate parameters $k_{\n{nup}}$ for
	influx and $\xi k_{\n{nup}}$ for efflux.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/VaryingC.eps}
		\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
			All parameters but $\n{[c]_a}$ are as presented in
			Table~\ref{table1:nondim}. The extracellular xanthosine
			concentration in these plots is $\n{[c]_a}=10^{1}$ in (A) and
			$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall that $\n{[c]_{a}} \defeq
			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
			so $\n{[c]_a}$ is dimensionless).
			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline) up
			or down, but the blue curve (mRNA nullcline) is unchanged (see
			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
			only the lower fixed point (fixed point number 1), whereas in (B)
			there is only the upper one (fixed point number 3). In between lies
			the bistable case that was shown in Fig~\ref{fig4:bistability}.}
		\label{fig5:extraxanth}
	\end{figure}

@@

	\paragraph*{The roles of XapA and XapB.}
	While it is clear that the bistability in the model system is due to the
	feedback loop from XapA and XapB, it is not obvious if both XapA and XapB
	are necessary. It turns out that the bistability is due to XapB only. XapA
	is neither sufficient nor necessary and, within the estimated parameter
	regime, does not even have a significant influence on the system. This can
	be seen from the plots in figure~\ref{fig6:xapAB}.
	Degradation of xanthosine by XapA lowers the xanthosine and protein concentration at the upper
	fixed point by a small amount and could, in principle, thereby make the
	high-expression solution vanish. For our choice of all other parameters,
	bistability only vanishes for $k_{\n{\alpha}} > 10^4$ which is far from what has been
	measured. However, a higher effective rate could, in principle, be achieved by
	different translation rates of XapA and XapB (see simplifications of the
	model in~\nameref{S1_Text}). Hence, we cannot exclude the possibility that
	XapA becomes so strong that it makes bistability impossible, but this is an
	extreme case. XapB, on the other hand, is essential; without it the system
	only has the one fixed point at low expression.\footnote{
		Seeger et.\ al.\cite{Seeger1995} observed that $\Delta xapB$ mutants could
		survive, but grew extremely slowly, with xanthosine as the only carbon
		source, which makes sense in light of Fig~\ref{fig6:xapAB}.
		With \textit{xapB} removed, the switch never activates and the cells are
		forced to survive with an extremely meager quantity of XapA to metabolize
		the abundant xanthosine.}

@@

	For a cell, the minimal effect of XapA on bistability is a useful feature: by coupling XapA and XapB on an
	operon, XapA is switched on and off together with XapB but it does not
	significantly disturb this adaptation mechanism, while its kinetic
	parameters and expression levels can be chosen 
	somewhat freely as necessary for metabolism. By
	having a membrane transporter gene on an operon whose expression is
	activated by the transporter substrate, the expression of a whole set of
	enzymes can be turned on and off depending on the presence of the substrate.
	It seems likely that this mechanism of short-term adaptation of a single cell to
	its environment may be used by cells for many metabolic processes.

@@

	If, as a purely theoretical consideration, we remove either the second
	xanthosine binding site on XapR or the second XapR binding site on the
	promoter, leaving cooperativity in only one component of the system,
	we find that the system still has a bistable parameter regime.
	However, this bistable parameter range is smaller than in the original
	model, which makes the system less stable: small stochastic fluctuations
	in the parameter values can collapse the system to monostability,
	possibly leaving it in the wrong state and without its ability to adapt.
	But only when the second binding site is removed in both places, leaving
	no cooperativity in the system, are the
	equations are insufficiently non-linear to produce bistability. An example
	of the three scenarios (only cooperative XapR, only cooperative promoter, no cooperativity) can be seen in Fig~\ref{fig7:coop}. It follows that
	there need to be either two xanthosine binding sites on XapR or two XapR
	binding sites on the promoter (or both) in order to obtain a switch-like
	behavior.

@@

	One can also ask how much cooperative interaction is needed between the two
	binding sites. For the promoter, the amount of cooperativity is given by $\Delta E_{\n{coop}}$ in
	our model, and we find that setting $\Delta E_{\n{coop}} = 0$ has almost no
	influence on the phase diagrams. For XapR, we cannot test how much interaction is needed: the two binding sites interact
	indirectly, because the active state is much likelier if two xanthosine
	molecules are bound, and thus there is no continuous tuning parameter for the cooperative interaction
	like $\Delta E_{\n{coop}}$ in the case of the promoter.

@@

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations.

@@

		\bibitem{Wong1997}
		Wong P, Gladney S, Keasling JD.
		\newblock Mathematical Model of the \textit{lac} Operon: Inducer Exclusion,
		Catabolite Repression, and Diauxic Growth on Glucose and Lactose.
		\newblock Biotechnology Progress. 1997;13(2):132 -- 143.
		\newblock doi:{https://doi.org/10.1021/bp970003o}.
